---
title: "Data Vizualization Final Project Exploratory Analysis"
author: "Andrew Thvedt"
date: "April 25, 2020"
output: html_notebook
---
```{r}
#load required packages
library(tidyverse)

```

```{r, message = FALSE, warning = FALSE}
#read data
df <- read_csv('Data/yellow_tripdata_2019-10.csv')

#preview data
head(df, 10)
```

```{r}
#total missingness in data
sum(is.na(df))
```
233615 missing observations in the data for october.

```{r}
#missingness by column
colMeans(is.na(df))*100
```

However, the pickup time, dropoff time, pickup location, and dropoff location columns have 0 missing data. Most of the missingness is in the VendorID, store_and_fwd_flag, passenger_count, payment_type, and RatecodeID columns. This is not a big issue for our project, as these columns are less useful for us.

## Notes Moving Forward
* download and merge the TLC data containing pickup and dropoff zone IDs with current df
* google API to get long/lat data based on pickup zones
* map pickup zones (leaflet?) for visual representation

# Data Processing

```{r}
#read in jan-jun yellow taxi data

#january
jan <- read_csv('Data/yellow_tripdata_2019-01.csv')

#february
feb  <- read_csv('Data/yellow_tripdata_2019-02.csv')

#march
mar <- read_csv('Data/yellow_tripdata_2019-03.csv')

#april
apr <- read_csv('Data/yellow_tripdata_2019-04.csv')

#may
may <- read_csv('Data/yellow_tripdata_2019-05.csv')

#june
jun <- read_csv('Data/yellow_tripdata_2019-06.csv')

```


```{r}
#remove NA
jan <- na.omit(jan)
feb <- na.omit(feb)
mar <- na.omit(mar)
apr <- na.omit(apr)
may <- na.omit(may)
jun <- na.omit(jun)
```

